Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-21 aistats.tex
Add table of content for appendix only
Make norm matrices consistent
minor fix

@@

@@

@@

This bound is asymptotically tight.

@@

@@

\paragraph{Log-partition is Self-Concordant}
Adopting the notation of definition~\ref{def:self-concordance}, let us set $w=(\nat_1 \ 0)^\top$ and $v=(0 \ -1)^\top$.
Then 
\alignn{g(t) := 2A(w+tv) = \frac{a}{t} - \log(t)}
is well defined for $t>0$ and $a=\frac{\nat_1^2}{2} \geq 0$.
The logarithm is self-concordant for the same reason that the entropy is, but the inverse function $\inv{t}$ is not \citep[example 4.1.1.5]{nesterov2003introductory}.
Linear combinations with positive coefficients  preserves self-concordance \citep[theorem 4.1.1]{nesterov2003introductory} -- e.g. self-concordance is a cone, so it seems unlikely that this sum of a self-concordant function and a not self-concordant function would be self-concordant. 
But let us verify it to be sure. The derivatives of $g$ are
\alignn{
g'(t) &= -\frac{a}{t^2} - \frac{1}{t} \; ,\\
g''(t) &= \frac{2a}{t^3} + \inv{t^2} = \frac{1}{t^2}\paren{\frac{2a}{t} + 1} \; , \\
g'''(t) &= -\frac{6a}{t^4} - \frac{2}{t^3} = \frac{-2}{t^3}\paren{\frac{3a}{t} + 1} \;.
}
Then the definition of self-concordance reads
\alignn{
\abs{g'''(t)} &\leq 2 g''(t)^{\frac{3}{2}} \\
\iff \frac{3a}{t} + 1 &\leq \paren{\frac{2a}{t} + 1}^{\frac{3}{2}} \; .
}
Writing $y = \frac{a}{t} \geq 0$, and taking the square of this expression yields
\alignn{
 (3y + 1)^2 &\leq (2y + 1)^3 \\
 \iff 9 y ^2 + 6y + 1 &\leq 8 y^3  + 12 y^2 + 6y +1 \\
\iff 0 &\leq y^2(8y + 3)
}
which is true since $y\geq 0$.

The log-partition is neither self concordant nor pseudo-self-concordant.

\section{ASYMPTOTIC DERIVATION}
\label{app:asymptote}
In this section we fill-in the lines of \S\ref{ssec:asymptote} to prove Equation~\eqref{eq:asymptote}.
Approximating $\conj$ with a second order Taylor expansion yields
\aligns{
    \bregmanconj(\m^* ; \m)
    &= \frac{1}{2}\norm{\m - \m^*}^2_{\mF}
    + O(\norm{\m - \m^*}^3),
}
where the norm is induced by the matrix
\aligns{
    \mF
    :=\nabla^2\conj(\m^*)
    = \nabla^2\logpart(\nat^*)^{-1}
    = \Cov_{\nat^*}(T)^{-1} 
}
where the second equality is a general property of convex conjugates.
Plugging the MLE~\eqref{eq:defMLE} into this quadratic and expanding it yields
\aligns{
	\E \half \norm{\inv{n}  \sum_i T_i - \m^* }_\mF^2 
	&=\inv{2n^2} \sum_i \E \norm{T_i - \m^* }_\mF^2 
	+ \inv{2n^2} \sum_{i\neq j}\expect{T_i - \m^*}^\top \mF \overbrace{\expect{T_j - \m^*}}^{0} ]\\
	&=\inv{2n}  \E \norm{T_1 - \m^* }_\mF^2 \\
	& = \inv{2n} \Tr(\mF\ \expect{(T_1 - \m^*) (T_1 - \m^*)^\top}) \\
	&= \inv{2 n} \Tr(\mF \Cov_{\nat^*}(T)) = \frac{d}{2n} 
}
where on the first line we used independence of samples, on the second line we used the fact that samples are identically distributed, and on the third line we used the  trace trick along with the linearity of the trace $\Tr$.
On the way, this also proves that for the MLE $\norm{\m - \m^*}^3 \in O(n^{-\half[3]})$.
This yield the final rate for the MLE
\begin{align}
	\E \bregmanconj \paren{\E [T(X)] ; \inv{n}  \sum_i T_i}
	= \frac{d}{2n} + O(n^{- \frac{3}{2}} ) \; .
\end{align}
For the MAP~\eqref{eq:defMAP}, the quadratic decomposes into bias and variance
\begin{align}
	\E \half \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
	&= \frac{n d}{2(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \half \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
	&= \frac{d}{2n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
\end{align}
This $O(n^{-2})$ term is dominated by the $O(n^{-\half[3]})$ term from the quadratic approximation of the Bregman, yielding the same first order rate as for the MLE
\begin{align}
	\E \bregmanconj \paren{\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }
	= \frac{d}{2n} + O(n^{- \frac{3}{2}} ) \; .
\end{align}

>>>>

% To make a table of contents only for the appendix
\usepackage[toc,page,header]{appendix}
\usepackage{minitoc}
\renewcommand \thepart{}
\renewcommand \partname{}

@@

\doparttoc % Tell to minitoc to generate a toc for the parts
\faketableofcontents % Run a fake tableofcontents command for the partocs
%\part{} % Start the document part
%\parttoc % Insert the document TOC

@@

\addcontentsline{toc}{section}{Appendix} % Add the appendix text to the document TOC
\part{Appendix} % Start the appendix part
\parttoc % Insert the appendix TOC

@@

@@

We see from~\eqref{eq:asymptote} that this bound is asymptotically tight.

@@

\paragraph{Log-partition is Self-Concordant}
Adopting the notation of definition~\ref{def:self-concordance}, let us set $w=(\nat_1 \ 0)^\top$ and $v=(0 \ -1)^\top$.
Then 
\alignn{g(t) := 2A(w+tv) = \frac{a}{t} - \log(t)}
is well defined for $t>0$ and $a=\frac{\nat_1^2}{2} \geq 0$.
The logarithm is self-concordant for the same reason that the entropy is, but the inverse function $\inv{t}$ is not \citep[example 4.1.1.5]{nesterov2003introductory}.
Linear combinations with positive coefficients  preserves self-concordance \citep[theorem 4.1.1]{nesterov2003introductory} -- e.g. self-concordance is a cone, so it seems unlikely that this sum of a self-concordant function and a not self-concordant function would be self-concordant. 
But let us verify it to be sure. The derivatives of $g$ are
\alignn{
g'(t) &= -\frac{a}{t^2} - \frac{1}{t} \; ,\\
g''(t) &= \frac{2a}{t^3} + \inv{t^2} = \frac{1}{t^2}\paren{\frac{2a}{t} + 1} \; , \\
g'''(t) &= -\frac{6a}{t^4} - \frac{2}{t^3} = \frac{-2}{t^3}\paren{\frac{3a}{t} + 1} \;.
}
Then the definition of self-concordance reads
\alignn{
\abs{g'''(t)} &\leq 2 g''(t)^{\frac{3}{2}} \\
\iff \frac{3a}{t} + 1 &\leq \paren{\frac{2a}{t} + 1}^{\frac{3}{2}} \; .
}
Writing $y = \frac{a}{t} \geq 0$, and taking the square of this expression yields
\alignn{
 (3y + 1)^2 &\leq (2y + 1)^3 \\
 \iff 9 y ^2 + 6y + 1 &\leq 8 y^3  + 12 y^2 + 6y +1 \\
\iff 0 &\leq y^2(8y + 3)
}
which is true since $y\geq 0$.
The log-partition is self concordant.

\section{ASYMPTOTIC DERIVATION}
\label{app:asymptote}
In this section we fill-in the lines of \S\ref{ssec:asymptote} to prove Equation~\eqref{eq:asymptote}.
Approximating $\conj$ with a second order Taylor expansion yields
\aligns{
    \bregmanconj(\m^* ; \m)
    &= \frac{1}{2}\norm{\m - \m^*}^2_{\mF}
    + O(\norm{\m - \m^*}^3),
}
where the norm is induced by the matrix
\aligns{
    \mF
    :=\nabla^2\conj(\m^*)
    = \nabla^2\logpart(\nat^*)^{-1}
    = \Cov_{\nat^*}(T)^{-1} 
}
where the second equality is a general property of convex conjugates.
Plugging the MLE~\eqref{eq:defMLE} into this quadratic and expanding it yields
\aligns{
	\E \half \norm{\inv{n}  \sum_i T_i - \m^* }_\mF^2 
	&=\inv{2n^2} \sum_i \E \norm{T_i - \m^* }_\mF^2 
	+ \inv{2n^2} \sum_{i\neq j}\expect{T_i - \m^*}^\top \mF \overbrace{\expect{T_j - \m^*}}^{0} ]\\
	&=\inv{2n}  \E \norm{T_1 - \m^* }_\mF^2 \\
	& = \inv{2n} \Tr(\mF\ \expect{(T_1 - \m^*) (T_1 - \m^*)^\top}) \\
	&= \inv{2 n} \Tr(\mF \Cov_{\nat^*}(T)) = \frac{d}{2n} 
}
where on the first line we used independence of samples, on the second line we used the fact that samples are identically distributed, and on the third line we used the  trace trick along with the linearity of the trace $\Tr$.
On the way, this also proves that for the MLE $\norm{\m - \m^*}^3 \in O(n^{-\half[3]})$.
This yield the final rate for the MLE
\begin{align}
	\E \bregmanconj \paren{\E [T(X)] ; \inv{n}  \sum_i T_i}
	= \frac{d}{2n} + O(n^{- \frac{3}{2}} ) \; .
\end{align}
For the MAP~\eqref{eq:defMAP}, the quadratic decomposes into bias and variance
\begin{align}
	\E \half \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mF}
	&= \frac{n d}{2(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \half \norm{\mu^* -  \mu_0}^2_{\mF} \\
	&= \frac{d}{2n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mF} }{n^2} \right) \; .
\end{align}
This $O(n^{-2})$ term is dominated by the $O(n^{-\half[3]})$ term from the quadratic approximation of the Bregman, yielding the same first order rate as for the MLE
\begin{align}
	\E \bregmanconj \paren{\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }
	= \frac{d}{2n} + O(n^{- \frac{3}{2}} ) \; .
\end{align}
